%  Copyright (C) 2002-2003 David Roundy
%
%  This program is free software; you can redistribute it and/or modify
%  it under the terms of the GNU General Public License as published by
%  the Free Software Foundation; either version 2, or (at your option)
%  any later version.
%
%  This program is distributed in the hope that it will be useful,
%  but WITHOUT ANY WARRANTY; without even the implied warranty of
%  MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
%  GNU General Public License for more details.
%
%  You should have received a copy of the GNU General Public License
%  along with this program; see the file COPYING.  If not, write to
%  the Free Software Foundation, Inc., 51 Franklin Street, Fifth Floor,
%  Boston, MA 02110-1301, USA.

\documentclass{book}
%\usepackage{color}

\usepackage{verbatim}
\usepackage{html}
\usepackage{fancyvrb}
\newenvironment{code}{\comment}{\endcomment}
% \newenvironment{code}{\color{blue}\verbatim}{\endverbatim}

\begin{document}


% Definition of title page:
\title{Darcs User Manual}
\date{
\darcsVersion
} % icky newline before closing brace is to appease preproc.hs.
\author{David Roundy}

\maketitle

\tableofcontents

\chapter{Introduction}

This manual provides a stable documentation for using darcs.
To find more up-to-date and complementary information, please consult the
darcs wiki at
\htmladdnormallinkfoot{http://wiki.darcs.net}{http://wiki.darcs.net}.

Darcs is a revision control system, along the lines of Subversion, Git or
Mercurial.  That means that it keeps track of various revisions and branches of your
project, allows for changes to propagate from one branch to another.  Darcs
is intended to be an ``advanced'' revision control system.  Darcs has two
particularly distinctive features which differ from other revision control
systems: 1) each copy of the source is a fully functional branch, and 2)
it is based on changes and not revisions.

\paragraph{Every source tree a branch}
The primary simplifying notion of darcs is that \emph{every} copy of your
source code is a full repository. This is dramatically different from Subversion,
in which the normal usage is for there to be one central repository from
which source code will be checked out. It is closer to the notion of arch,
since the `normal' use of arch is for each developer to create his own
repository. However, darcs makes it even easier, since simply checking out
the code is all it takes to create a new repository. This has several
advantages, since you can harness the full power of darcs in any scratch
copy of your code, without committing your possibly destabilizing changes to
a central repository.

%% FIXME: unbound personal pronoun (me).  Use third person, passive voice.

\paragraph{Change-oriented}
The development of a simplified theory of patches is what originally
motivated the creation of darcs. This patch formalism means that darcs patches
have a set of properties, which make possible manipulations that couldn't be
done in other revision control systems. First, every patch is invertible.
Secondly, sequential patches (i.e.\ patches that are created in sequence, one
after the other) can be reordered, although this reordering can fail, which
means the second patch is dependent on the first. Thirdly, patches which are
in parallel (i.e.\ both patches were created by modifying identical trees)
can be merged, and the result of a set of merges is independent of the order
in which the merges are performed. This last property is critical to darcs'
philosophy, as it means that a particular version of a source tree is fully
defined by the list of patches that are in it, i.e.\ there is no issue
regarding the order in which merges are performed. For a more thorough
discussion of darcs' theory of patches, see Appendix~\ref{Patch}.

%% FIXME: clarify - the theory is advanced, the UI is simple.

\paragraph{A simple advanced tool}
Besides being ``advanced'' as discussed above, darcs is actually also quite
simple. Versioning tools can be seen as three layers. At the foundation is
the ability to manipulate changes. On top of that must be placed some kind
of database system to keep track of the changes. Finally, at the very top is
some sort of distribution system for getting changes from one place to
another.

In darcs, only the first of these three layers is of particular interest,
so the other two are done as simply as possible.  At the database
layer, darcs just has an ordered list of patches along with the patches
themselves, each stored as an individual file.  Darcs' distribution system
is based on a dumb server, typically apache or just a local or network file
system when pulling patches.
darcs has built-in support for using \verb!ssh! to write to a remote file
system. A darcs executable is called on the remote system to apply the patches.
Arbitrary other transport protocols are supported, through an environment
variable describing a command that will run darcs on the remote system.
See the documentation for DARCS\_APPLY\_FOO in Chapter~\ref{configuring}
for details.

The recommended method is to send patches through gpg-signed email
messages, which has the advantage of being mostly asynchronous.

\paragraph{Keeping track of changes rather than versions}

In the last paragraph, I explained revision control systems in terms of
three layers.  One can also look at them as having two distinct uses.  One
is to provide a history of previous versions.  The other is to keep track
of changes that are made to the repository, and to allow these changes to
be merged and moved from one repository to another.  These two uses are
distinct, and almost orthogonal, in the sense that a tool can support one
of the two uses optimally while providing no support for the other.  Darcs
is not intended to maintain a history of versions, although it is possible
to kludge together such a revision history, either by making each new patch
depend on all previous patches, or by tagging regularly.  In a sense, this
is what the tag feature is for, but the intention is that tagging will be
used only to mark particularly notable versions (e.g.\ released versions, or
perhaps versions that pass a time consuming test suite).

Other revision control systems are centered upon the job of keeping track
of a history of versions, with the ability to merge changes being added as
it was seen that this would be desirable.  But the fundamental object
remained the versions themselves.

In such a system, a patch (I am using patch here to mean an encapsulated
set of changes) is uniquely determined by two trees.  Merging changes that
are in two trees consists of finding a common parent tree, computing the
diffs of each tree with their parent, and then cleverly combining those two
diffs and applying the combined diff to the parent tree, possibly at some
point in the process allowing human intervention, to allow for fixing up
problems in the merge such as conflicts.

In the world of darcs, the source tree is \emph{not} the fundamental
object, but rather the patch is the fundamental object.  Rather than a
patch being defined in terms of the difference between two trees, a tree is
defined as the result of applying a given set of patches to an empty tree.
Moreover, these patches may be reordered (unless there are dependencies
between the patches involved) without changing the tree.  As a result,
there is no need to find a common parent when performing a merge.  Or, if
you like, their common parent is defined by the set of common patches, and
may not correspond to any version in the version history.

One useful consequence of darcs' patch-oriented philosophy is that since a
patch need not be uniquely defined by a pair of trees (old and new), we can
have several ways of representing the same change, which differ only in how
they commute and what the result of merging them is.  Of course, creating
such a patch will require some sort of user input.  This is a Good Thing,
since the user \emph{creating} the patch should be the one forced to think
about what he really wants to change, rather than the users merging the
patch.  An example of this is the token replace patch (See
Section~\ref{token_replace}).  This feature makes it possible to create a
patch, for example, which changes every instance of the variable
``stupidly\_named\_var'' to ``better\_var\_name'', while leaving
``other\_stupidly\_named\_var'' untouched.  When this patch is merged with
any other patch involving the ``stupidly\_named\_var'', that instance will
also be modified to ``better\_var\_name''.  This is in contrast to a more
conventional merging method which would not only fail to change new
instances of the variable, but would also involve conflicts when merging
with any patch that modifies lines containing the variable.  By more using
additional information about the programmer's intent, darcs is thus able to
make the process of changing a variable name the trivial task that it
really is, which is really just a trivial search and replace, modulo
tokenizing the code appropriately.

The patch formalism discussed in Appendix~\ref{Patch} is what makes darcs'
approach possible.  In order for a tree to consist of a set of patches,
there must be a deterministic merge of any set of patches, regardless of the
order in which they must be merged.  This requires that one be able to
reorder patches.  While I don't know that the patches are required to be
invertible as well, my implementation certainly requires invertibility.  In
particular, invertibility is required to make use of
Theorem~\ref{merge_thm}, which is used extensively in the manipulation of
merges.

\input{features.tex}

\chapter{Getting started}

This chapter will lead you through an example use of darcs, which hopefully
will allow you to get started using darcs with your project.

\section{Creating your repository}

Creating your repository in the first place just involves telling darcs to
create the special directory (called {\tt \_darcs}) in your project tree,
which will hold the revision information.  This is done by simply calling
from the root directory of your project:
\begin{verbatim}
$ cd my_project/
$ darcs initialize
\end{verbatim}
This creates the \verb|_darcs| directory and populates it with whatever
files and directories are needed to describe an empty project.  You now
need to tell darcs what files and directories in your project should be
under revision control.  You do this using the command \verb|darcs add|:
\begin{verbatim}
$ darcs add *.c Makefile.am configure.ac
\end{verbatim}
When you have added all your files (or at least, think you have), you will
want to record your changes.  ``Recording'' always includes adding a note
as to why the change was made, or what it does.  In this case, we'll just
note that this is the initial version.
\begin{verbatim}
$ darcs record --all
What is the patch name? Initial revision.
\end{verbatim}
Note that since we didn't specify a patch name on the command line we were
prompted for one.  If the environment variable `EMAIL' isn't set, you will
also be prompted for your email address.  Each patch that is recorded is
given a unique identifier consisting of the patch name, its creator's email
address, and the date when it was created.

\section{Making changes}

Now that we have created our repository, make a change to one or more of
your files.  After making the modification run:
\begin{verbatim}
$ darcs whatsnew
\end{verbatim}
This should show you the modifications that you just made, in the darcs
patch format.  If you prefer to see your changes in a different format,
read Section~\ref{whatsnew}, which describes the whatsnew command in
detail.

Let's say you have now made a change to your project.  The next thing to do
is to record a patch.  Recording a patch consists of grouping together a
set of related changes, and giving them a name.  It also tags the patch
with the date it was recorded and your email address.

To record a patch simply type:
\begin{verbatim}
$ darcs record
\end{verbatim}
darcs will then prompt you with all the changes that you have made that
have not yet been recorded, asking you which ones you want to include in
the new patch.  Finally, darcs will ask you for a name for the patch.

You can now rerun whatsnew, and see that indeed the changes you have
recorded are no longer marked as new.

\section{Making your repository visible to others}
How do you let the world know about these wonderful changes?  Obviously,
they must be able to see your repository.  Currently the easiest way to do
this is typically by http using any web server.  The recommended way to do
this (using apache in a UNIX environment) is to create a directory called
{\tt /var/www/repos}, and then put a symlink to your repository there:
\begin{verbatim}
$ cd /var/www/repos
$ ln -s /home/username/myproject .
\end{verbatim}

\section{Getting changes made to another repository}
Ok, so I can now browse your repository using my web browser\ldots\ so
what? How do I get your changes into \emph{my} repository, where they can
do some good? It couldn't be easier.  I just \verb|cd| into my repository,
and there type:
\begin{verbatim}
$ darcs pull http://your.server.org/repos/yourproject
\end{verbatim}
Darcs will check to see if you have recorded any changes that aren't in my
current repository.  If so, it'll prompt me for each one, to see which ones
I want to add to my repository.  Note that you may see a different series
of prompts depending on your answers, since sometimes one patch depends on
another, so if you answer yes to the first one, you won't be prompted for
the second if the first depends on it.

Of course, maybe I don't even have a copy of your repository.  In that case
I'd want to do a
\begin{verbatim}
$ darcs get http://your.server.org/repos/yourproject
\end{verbatim}
which gets the whole repository.

I could instead create an empty repository and fetch all of your patches
with pull.  Get is just a more efficient way to clone a whole repository.

Get, pull and push also work over ssh.  Ssh-paths are of the same form
accepted by scp, namely \verb|[username@]host:/path/to/repository|.

\section{Moving patches from one repository to another}

Darcs is flexible as to how you move patches from one repository to another.
This section will introduce all the ways you can get patches from one place
to another, starting with the simplest and moving to the most complicated.

\subsection{All pulls}

The simplest method is the ``all-pull'' method.  This involves making each
repository readable (by http, ftp, nfs-mounted disk, whatever), and you
run \verb|darcs pull| in the repository you want to move the patch to.  This is nice,
as it doesn't require you to give write access to anyone else, and is
reasonably simple.

\subsection{Send and apply manually}

Sometimes you have a machine on which it is not convenient to set up a web
server, perhaps because it's behind a firewall or perhaps for security
reasons, or because it is often turned off.  In this case you can use
\verb|darcs send|
from that computer to generate a patch bundle destined for another
repository.  You can either let darcs email the patch for you, or save it
as a file and transfer it by hand.  Then in the destination repository you
(or the owner of that repository) run \verb|darcs apply| to apply the patches contained
in the bundle.  This is also quite a simple method since, like the all-pull
method, it doesn't require that you give anyone write access to your
repository.  But it's less convenient, since you have to keep track of the
patch bundle (in the email, or whatever).

If you use the send and apply method with email, you'll probably want to
create a \verb!_darcs/prefs/email! file containing your email address.
This way anyone who sends to your repository will automatically send the
patch bundle to your email address.

If you receive many patches by email, you probably will benefit by running
darcs apply directly from your mail program.  I have in my \verb!.muttrc!
the following:
\begin{verbatim}
auto_view text/x-patch text/x-darcs-patch
macro pager A "<pipe-entry>darcs apply --verbose --mark-conflicts \
        --reply droundy@abridgegame.org --repodir ~/darcs"
\end{verbatim}
which allows me to view a sent patch, and then apply the patch directly from \verb!mutt!, sending a
confirmation email to the person who sent me the patch. The autoview line relies on on the following
lines, or something like them, being present in one's \verb!.mailcap!:
\begin{verbatim}
text/x-patch;                           cat; copiousoutput
text/x-darcs-patch;                     cat; copiousoutput
\end{verbatim}

\subsection{Push}

If you use ssh (and preferably also ssh-agent, so you won't have to keep
retyping your password), you can use the push method to transfer changes
(using the scp protocol for communication).  This method is again not very
complicated, since you presumably already have the ssh permissions set up.
Push can also be used when the target repository is local, in which case
ssh isn't needed.  On the other hand, in this situation you could as easily
run a pull, so there isn't much benefit.

Note that you can use push to administer a multiple-user repository.  You
just need to create a user for the repository (or repositories), and give
everyone with write access ssh access, perhaps using
\verb!.ssh/authorized_keys!.  Then they run
\begin{verbatim}
$ darcs push repouser@repo.server:repo/directory
\end{verbatim}

\subsection{Push ---apply-as}

Now we get more subtle.  If you like the idea in the previous paragraph
about creating a repository user to own a repository which is writable by
a number of users, you have one other option.

Push \verb!--apply-as! can run on either a local repository or one accessed
with ssh, but uses \verb!sudo! to run a darcs apply command (having created
a patch bundle as in send) as another user.  You can add the following line
in your \verb|sudoers| file to allow the users to apply their patches to a
centralized repository:
{\small
\begin{verbatim}
ALL   ALL = (repo-user) NOPASSWD: /usr/bin/darcs apply --all --repodir /repo/path*
\end{verbatim}
}
This method is ideal for a centralized repository when all the users have
accounts on the same computer, if you don't want your users to be able to
run arbitrary commands as repo-user.

\subsection{Sending signed patches by email}

Most of the previous methods are a bit clumsy if you don't want to give
each person with write access to a repository an account on your server.  Darcs
send can be configured to send a cryptographically signed patch by email.
You can then set up your mail system to have darcs verify that patches were
signed by an authorized user and apply them when a patch is received by
email.  The results of the apply can be returned to the user by email.
Unsigned patches (or patches signed by unauthorized users) will be
forwarded to the repository owner (or whoever you configure them to be
forwarded to\ldots).

This method is especially nice when combined with the \verb!--test! option
of darcs apply, since it allows you to run the test suite (assuming you
have one) and reject patches that fail---and it's all done on the server,
so you can happily go on working on your development machine without
slowdown while the server runs the tests.

Setting up darcs to run automatically in response to email is by far the
most complicated way to get patches from one repository to another\ldots\ so it'll
take a few sections to explain how to go about it.

\paragraph{Security considerations}

When you set up darcs to run apply on signed patches, you should assume
that a user with write access can write to any file or directory that is
writable by the user under which the apply process runs.  Unless you
specify the \verb!--no-test! flag to darcs apply (and this is \emph{not}
the default), you are also allowing anyone with write access to that
repository to run arbitrary code on your machine (since they can run a test
suite---which they can modify however they like).  This is quite a
potential security hole.

For these reasons, if you don't implicitly trust your users, it is
recommended that you create a user for each repository to limit the damage
an attacker can do with access to your repository.  When considering who to
trust, keep in mind that a security breach on any developer's machine could
give an attacker access to their private key and passphrase, and thus to
your repository.

\paragraph{Installing necessary programs}

You also must install the following programs: gnupg, a mailer configured to
receive mail (e.g.\ exim, sendmail or postfix), and a web server (usually
apache).

\paragraph{Granting access to a repository}

You create your gpg key by running (as your normal user):
\begin{verbatim}
$ gpg --gen-key
\end{verbatim}
You will be prompted for your name and email address, among other options.
%%To add your public key to the allowed keys keyring.
Of course, you can
skip this step if you already have a gpg key you wish to use.

You now need to export the public key so we can tell the patcher about it.
You can do this with the following command (again as your normal user):
\begin{verbatim}
$ gpg --export "email@address" > /tmp/exported_key
\end{verbatim}
And now we can add your key to the \verb!allowed_keys!:
\begin{verbatim}
(as root)> gpg --keyring /var/lib/darcs/repos/myproject/allowed_keys \
               --no-default-keyring --import /tmp/exported_key
\end{verbatim}
You can repeat this process any number of times to authorize multiple users
to send patches to the repository.

You should now be able to send a patch to the repository by running as your
normal user, in a working copy of the repository:
\begin{verbatim}
$ darcs send --sign http://your.computer/repos/myproject
\end{verbatim}
You may want to add ``send sign'' to the file \verb!_darcs/prefs/defaults!
so that you won't need to type \verb!--sign! every time you want to
send\ldots

If your gpg key is protected by a passphrase, then executing \verb!send!
with the \verb!--sign! option might give you the following error:
\begin{verbatim}
darcs failed:  Error running external program 'gpg'
\end{verbatim}
The most likely cause of this error is that you have a misconfigured
gpg that tries to automatically use a non-existent gpg-agent
program. GnuPG will still work without gpg-agent when you try to sign
or encrypt your data with a passphrase protected key. However, it will
exit with an error code 2 (\verb!ENOENT!) causing \verb!darcs! to
fail. To fix this, you will need to edit your \verb!~/.gnupg/gpg.conf!
file and comment out or remove the line that says:
\begin{verbatim}
use-agent
\end{verbatim}
If after commenting out or removing the \verb!use-agent! line in your
gpg configuration file you still get the same error, then you probably
have a modified GnuPG with use-agent as a hard-coded option. In that
case, you should change \verb!use-agent! to \verb!no-use-agent! to
disable it explicitly.

\paragraph{Setting up a sendable repository using procmail}
If you don't have root access on your machine, or perhaps simply don't want
to bother creating a separate user, you can set up a darcs repository using
procmail to filter your mail.  I will assume that you already use procmail
to filter your email.  If not, you will need to read up on it, or perhaps
should use a different method for routing the email to darcs.

To begin with, you must configure your repository so that a darcs send to
your repository will know where to send the email.  Do this by creating a
file in \verb!/path/to/your/repo/_darcs/prefs! called \verb!email!
containing your email address.  As a trick (to be explained below), we will
create the email address with ``darcs repo'' as your name, in an email
address of the form ``David Roundy $<$droundy@abridgegame.org$>$.''
\begin{verbatim}
$ echo 'my darcs repo <user@host.com>' \
      > /path/to/your/repo/_darcs/prefs/email
\end{verbatim}

The next step is to set up a gnupg keyring containing the public keys of
people authorized to send to your repository.  Here I'll give a second way of
going about this (see above for the first).  This time I'll assume you
want to give me write access to your repository.  You can do this by:
\begin{verbatim}
gpg --no-default-keyring \
    --keyring /path/to/the/allowed_keys --recv-keys D3D5BCEC
\end{verbatim}
This works because ``D3D5BCEC'' is the ID of my gpg key, and I have
uploaded my key to the gpg keyservers.  Actually, this also requires that
you have configured gpg to access a valid keyserver.  You can, of course,
repeat this command for all keys you want to allow access to.

Finally, we add a few lines to your \verb!.procmailrc!:
\begin{verbatim}
:0
* ^TOmy darcs repo
|(umask 022; darcs apply --reply user@host.com \
    --repodir /path/to/your/repo --verify /path/to/the/allowed_keys)
\end{verbatim}
The purpose for the ``my darcs repo'' trick is partially to make it easier
to recognize patches sent to the repository, but is even more crucial to
avoid nasty bounce loops by making the \verb!--reply! option have an email
address that won't go back to the repository.  This means that unsigned
patches that are sent to your repository will be forwarded to your ordinary
email.

Like most mail-processing programs, Procmail by default sets a tight umask.
However, this will prevent the repository from remaining world-readable;
thus, the ``umask 022'' is required to relax the umask.
(Alternatively, you could set Procmail's global \verb!UMASK! variable
to a more suitable value.)

\paragraph{Checking if your e-mail patch was applied}

After sending a patch with \verb!darcs send!, you may not receive any feedback,
even if the patch is applied. You can confirm whether or not your patch was applied
to the remote repository by pointing \verb!darcs changes! at a remote repository:
\begin{verbatim}
darcs changes --last=10 --repo=http://darcs.net/
\end{verbatim}

That shows you the last 10 changes in the remote repository. You can adjust the options given
to \verb!changes! if a more advanced query is needed.

\input{configuring_darcs.tex}

\input{best_practices.tex}

\input{formats.tex}

\chapter{Darcs commands}

\input{Darcs/Commands.lhs}

\section{Options apart from darcs commands}
\begin{options}
--help
\end{options}
Calling darcs with just \verb|--help| as an argument gives a brief
summary of what commands are available.
\begin{options}
--version, --exact-version
\end{options}
Calling darcs with the flag \verb|--version| tells you the version of
darcs you are using.  Calling darcs with the flag \verb|--exact-version|
gives the precise version of darcs, even if that version doesn't correspond
to a released version number.  This is helpful with bug reports, especially
when running with a ``latest'' version of darcs.
\begin{options}
--commands
\end{options}
Similarly calling darcs with only \verb|--commands| gives a simple list
of available commands.  This latter arrangement is primarily intended for
the use of command-line autocompletion facilities, as are available in
bash.

\section{Getting help}

\input{Darcs/Commands/Help.lhs}

\section{Creating repositories}

\input{Darcs/Commands/Init.lhs}

\input{Darcs/Commands/Get.lhs}

\input{Darcs/Commands/Put.lhs}

\section{Modifying the contents of a repository}

\input{Darcs/Commands/Add.lhs}

\input{Darcs/Commands/Remove.lhs}

\input{Darcs/Commands/Move.lhs}

\input{Darcs/Commands/Replace.lhs}

\section{Working with changes}

\input{Darcs/Commands/Record.lhs}

\input{Darcs/Commands/Pull.lhs}

\input{Darcs/Commands/Push.lhs}

\input{Darcs/Commands/Send.lhs}

\input{Darcs/Commands/Apply.lhs}

\section{Seeing what you've done}

\input{Darcs/Commands/WhatsNew.lhs}

\input{Darcs/Commands/Changes.lhs}

\input{Darcs/Commands/Show.lhs}

\section{More advanced commands}

\input{Darcs/Commands/Tag.lhs}

\input{Darcs/Commands/SetPref.lhs}

\input{Darcs/Commands/Check.lhs}

\input{Darcs/Commands/Optimize.lhs}

\section{Undoing, redoing and running in circles}

\input{Darcs/Commands/AmendRecord.lhs}

\input{Darcs/Commands/Rollback.lhs}

\input{Darcs/Commands/Unrecord.lhs}

\input{Darcs/Commands/Revert.lhs}

\input{Darcs/Commands/Unrevert.lhs}

\section{Advanced examination of the repository}

\input{Darcs/Commands/Diff.lhs}

\input{Darcs/Commands/Annotate.lhs}

% Includes the show commands.
\input{Darcs/Commands/Show.lhs}

\section{Rarely needed and obscure commands}

\input{Darcs/Commands/Convert.lhs}

\input{Darcs/Commands/MarkConflicts.lhs}

\input{Darcs/Commands/Dist.lhs}

\input{Darcs/Commands/TrackDown.lhs}

\input{Darcs/Commands/Repair.lhs}

\appendix

\input{building_darcs.tex}

\input{Darcs/Patch.lhs}

\input{Darcs/Repository/DarcsRepo.lhs}

\input{gpl.tex}

\end{document}


